\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-03-05 Paper/paper.tex
bitslicing perm and sub

The permutation $P$ works on individual bits and maps bit $b_i$ to $b_{P(i)},
\forall i\in\{0,1,\dots,n-1\}$. The different permutations for \texttt{GIFT-64}
and \texttt{GIFT-128} can be expressed by:

@@

We will take a look at NEON as this is the type of vector processing supported
by the Cortex A-73 processor.

@@

Bitslicing refers to the technique of splitting up $n$ bits into $m$ slices to
achieve a more efficient representation to operate on. The structure of \texttt{GIFT}
naturally offers possibilities for bitslicing. We split the cipher state bits $b_{63}b_{62}\dots b_0$
into four slices $S_i,0\leq i\leq 3$:

@@

>>>>

The permutation $P$ works on individual bits and maps bit $b_i$ to $b_{P(i)},
i\in\{0,1,\dots,n-1\}$. The different permutations for \texttt{GIFT-64} and
\texttt{GIFT-128} can be expressed by:

@@

We will take a look at NEON as this is the type of vector processing supported
by the Cortex-A73 processor.

@@

Bitslicing refers to the technique of splitting up $n$ bits into $m$ slices to
achieve a more efficient representation to operate on. The structure of
\texttt{GIFT} naturally offers possibilities for bitslicing. We split the
cipher state bits $b_{63}b_{62}\dots b_0$ into four slices $S_i,
i\in\{0,1,2,3\}$ such that the $i$-th slice contains all $i$-th bits of the
individual S-boxes.

@@

This representation offers multiple advantages. We first note that computation
of the S-box can be executed in parallel, similar to the \texttt{vperm}
technique above. This can be done by finding an algorithmic way to apply the
S-box which has already been proposed by the original \texttt{GIFT} authors:

\begin{align*}
    S_1&\leftarrow S_1\oplus (S_0\land S_2) \\
    t&\leftarrow S_0\oplus (S_1\land S_3) \\
    S_2&\leftarrow S_2\oplus (t\lor S_1) \\
    S_0&\leftarrow S_3\oplus S_2 \\
    S_1&\leftarrow S_1\oplus S_0 \\
    S_0&\leftarrow \lnot S_0 \\
    S_2&\leftarrow S_2\oplus (t\land S_1) \\
    S_3&\leftarrow t
\end{align*}

This is very efficient as it only requires six XOR-, three AND and one OR
operation.

An important property of the permutation is the fact that bits always stay in
their slice. This means we can decompose the permutation $P$ into four
permutations $P_i,i\in\{0,1,2,3\}$ and apply these permutations
separately to each slice. One possible way to implement a permutation $P_i$ in
software is to mask off all bits individually, shift them to their correct
position and OR them together:

\[
    P_i(S_i)=\bigvee_{k=0}^{15}{(S_i\land m_i) \ll s_i}
\]

This approach requires $47$ operations, meaning all four permutations require
over $150$ operations which would present a major bottleneck to the round
function. We can improve on this by working on multiple message blocks at once
and using the aforementioned \texttt{vperm} instruction to implement the bit
shuffling. We then need only four \texttt{TBL} instructions for the complete
diffusion layer.

\subsubsection{Using \texttt{vperm} for slice permutation}

We cannot use the \texttt{TBL} instruction directly as we need to shuffle
individual bits, but the smallest data we can operate on are bytes. We
therefore encrypt $8n$ messages at once which allows us to create bytewise
groupings. These messages are put into $4m$ registers with register $R_{4i}$
containing $S_0$, register $R_{4i+1}$ containing $S_1$ and so forth. With block
size $BS$ and register size $RS$, the following must hold:

\[
    8n\cdot BS=4m\cdot RS
\]

In the case of \texttt{GIFT-64} with $BS=64$ and ARM NEON with $RS=128$, we get

\[
    8n\cdot 64=4m\cdot 128\Leftrightarrow n=m
\]

We choose $n=m=1$, so for each encryption, we divide eight messages into four
registers.

\subsubsection{Packing the data into bitslice format}

Let $a,b,c,d,e,f,g,h$ be eight messages with subscripts denoting individual
bits. We first put these messages into four SIMD registers $V_0,V_1,V_2,V_3$:

\begin{align*}
    V_0&=b||a \\
    V_1&=d||c \\
    V_2&=f||e \\
    V_3&=h||g
\end{align*}

We then use the \texttt{SWAPMOVE} technique to bring the data into bitslice
format. This operation operates on two registers $A,B$ using mask $M$ and shift
value $N$. It swaps bits in $A$ masked by $M$ with bits masked by $M\ll N$ in
B using only three XOR-, one AND- and two shift operations.

\begin{align*}
    &\text{SWAPMOVE}(A,B,M,N): \\
    &\qquad T=((A\gg N)\oplus B)\land M \\
    &\qquad B=B\oplus T \\
    &\qquad A=A\oplus (T\ll N) \\
\end{align*}

The following operations put all of the first bits of each S-Box into $V_0$,
each second bit into $V_1$ and so forth and regroups these bits into bytes.

\begin{align*}
    &\text{SWAPMOVE}(V_0,V_1,0x5555\dots 55,1) &\text{SWAPMOVE}(V_0,V_0,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_2,V_3,0x5555\dots 55,1) &\text{SWAPMOVE}(V_1,V_1,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_0,V_2,0x3333\dots 33,2) &\text{SWAPMOVE}(V_2,V_2,0xf0f0\dots 0f,4) \\
    &\text{SWAPMOVE}(V_1,V_3,0x3333\dots 33,2) &\text{SWAPMOVE}(V_3,V_3,0xf0f0\dots 0f,4)
\end{align*}

With numbers denoting byte groups (i.e. $3$ corresponding to
$h_3g_3f_3e_3d_3c_3b_3a_3$), our data now has the following
permutation-friendly format:

\begin{alignat*}{7}
    V_0&=&&124&&||\dots&&||12&&||8 &&||4&&||0 \\
    V_1&=&&125&&||\dots&&||13&&||9 &&||5&&||1 \\
    V_2&=&&126&&||\dots&&||14&&||10&&||6&&||2 \\
    V_3&=&&127&&||\dots&&||15&&||11&&||7&&||3 \\
\end{alignat*}

We can now create permutation tables using the specification of the individual
slice permutations $P_i$:

\[
    \begin{array}{c|llllllllllllllll}
        j & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 \\
        \hline
        P_0(j) & 0 & 12 & 8 & 4 & 1 & 13 & 9 & 5 & 2 & 14 & 10 & 6 & 3 & 15 & 11 & 7 \\
        P_1(j) & 4 & 0 & 12 & 8 & 5 & 1 & 13 & 9 & 6 & 2 & 14 & 10 & 7 & 3 & 15 & 11 \\
        P_2(j) & 8 & 4 & 0 & 12 & 9 & 5 & 1 & 13 & 10 & 6 & 2 & 14 & 11 & 7 & 3 & 15 \\
        P_3(j) & 12 & 8 & 4 & 0 & 13 & 9 & 5 & 1 & 14 & 10 & 6 & 2 & 15 & 11 & 7 & 3
    \end{array}
\]

One thing to take note of is the original permutation values only showing where
a given byte should land, not which byte belongs to a certain position - i.e.
for $P_0$, byte 1 should land in position 12, but the byte belonging to
position 1 is byte 4. Because \texttt{vperm} works in the latter way, we have
to do some trivial rearrangements. Putting these values into registers
$V_5,V_6,V_7,V_8$ now allows us to compute the permutation layer in only four
instructions:

\begin{align*}
    &\texttt{TBL V0, V0, V5} &\texttt{TBL V1, V1, V6} \\
    &\texttt{TBL V2, V2, V7} &\texttt{TBL V3, V3, V8}
\end{align*}
